Galaxy Zoo: Morphological Classifications for 150,000 Galaxies in HST Legacy Imaging

### 2016-05-15 writeup/gz_hubble_data.tex
User -> classifier. Closes #93.

% Note: 250-word limit
We present the data release paper for the Galaxy Zoo: Hubble (GZH)
project.\footnote{This publication has been made possible by the participation
of more than 200,000 volunteers in the Galaxy Zoo project. Their contributions
are individually acknowledged at
\url{http://authors.galaxyzoo.org/authors.html}.} This is the third phase in a
large effort to measure reliable, detailed morphologies of galaxies by using
crowdsourced visual classifications of color composite images. Images in GZH
were selected from various \textit{Hubble Space Telescope} Legacy programs that
used the Advanced Camera for Surveys (AEGIS, COSMOS, GEMS, GOODS-N, and
GOODS-S), with filters selected to probe the rest-frame optical emission from
galaxies out to $z\simeq1$. The galaxies selected for GZH classifications go
down to magnitude limits of $m_{I814W}<23.5$ and have a median redshift of
$\langle z\rangle=0.9\pm0.6$, with a tail extending out to $z\simeq4$. The GZH
morphological data include measurements of both bulge- and disk-dominated
galaxies, details on spiral disk structure that relate to the Hubble type, as
well as clump identification and geometry. This paper also describes and
employs a novel method for calibrating morphological bias by using
artificially-redshifted galaxy images as a baseline. The GZH catalog contains
both raw and calibrated morphological vote fractions for $150,771$ galaxies,
providing a comprehensive data set suitable for large-scale studies of galaxy
evolution. 
% didn't the GOODS-N and GOODS-S second phases go deeper than 23.5 mag? am I misremembering? --BDS

@@

\subsection{User weighting}\label{ssec:weighting}

The votes of individual users who classified galaxies in GZH were combined to
make a vote fraction for each response ($f_{response}$) to a question in the
classification tree. Users' votes were weighted \citep[in a method identical to
that described in][]{wil13} such that users who frequently disagreed with all
other users ended up having very low weights. The user weighting $(w)$ was 1
for the top 95\% of users, as ranked by consistency. For the bottom 5\% of
users, $w$ was designed to drop smoothly and was effectively zero for the
bottom 1\% of classifiers.  Since this only affected a tiny percentage of
the users (and an even smaller percentage of the classifications) the
overall effect on the GZH dataset was minimal. The method was effective,
however, at filtering out contributions from random or deliberately
malicious classifiers.

Classifications each user were aggregated and weighted only if they
were logged into GZH under their username (which was encouraged, but not
required for participation). Classifications by users not logged in were marked
as ``Anonymous''; no weighting was applied to these classifications, but they
were otherwise treated identically in the data processing.

@@

\begin{figure*}
\center
\includegraphics[width=160mm]{figures/gzh_interface.png}
\caption{Screenshot of the GZH interface (\url{http:/zoo3.galaxyzoo.org}) at
the beginning of a classification, with the user ready to select an answer for
the first question in the decision tree.\label{fig:interface}}
\end{figure*}

Classifications for GZH were made using a web-based interface
(Figure~\ref{fig:interface}), similar in design to Galaxy~Zoo and Galaxy~Zoo~2.
The front-end runs on a Ruby~on~Rails framework with classifications stored in
a MySQL backend. Users were shown a randomly-selected color composite image from
the GZH sample; the default showed the image with a black sky background,
although users had the option to invert the color palette if desired. The
questions and responses for morphology appeared on the right side of the image as
a panel, including both text and icons. There was no tutorial required for
participation, although users could access an extensive ``Help'' section
containing example images and descriptive text for all the morphological
labels.

\begin{figure*}
\center
\includegraphics[width=\textwidth]{figures/gzh_decision_tree.pdf}
\caption{Flowchart of the questions presented to GZH users, labeled with the
corresponding Task numbers. Tasks in the decision tree are color-coded by tier
level: Gray-colored Tasks are 1$^\mathrm{st}$-tier questions which are asked of
all users. Tasks colored green, blue, purple, and pink (respectively) are one,
two, three, or four steps below branching points in the decision tree.}
\label{fig:decisiontree}
\end{figure*}

The procedure for classifying an image in GZH followed a hierarchical decision
tree (Figure~\ref{fig:decisiontree}). Every classification began with the user
identifying whether the object at the center of the image was a smooth galaxy,
a galaxy with a disk or other features, or a star/artifact. Subsequent
questions in the tree depended on the previous answer(s) given by the user; the
decision tree as designed so that the user answered every question relevant to
the morphology in the process of being identified. Questions that were not
answered were implicitly assumed to be absent in the image --- for example, if
the user identified a galaxy as being smooth, they were not asked to count the
number of spiral arms. For every task, the user chose a single answer before
continuing to the next question; they also had the option to restart any
classification in-progress.

The GZH decision tree had four broad sets of morphologies classified by the
users. The first option designated stars or image artifacts
(the result of either bad data or incorrect classification by the ACS pipeline
as a galaxy); in this case, the classification process ended and no further questions
were asked of the user. The second set were smooth galaxies, intended to select
ellipticals/early-types; users also measure the relative axial ratio
(roundness) for these galaxies. The third set was for disk/late-type galaxies,
which measured the features necessary to place a galaxy on the standard Hubble
tuning fork (bars, spiral arm, strength of the central bulge). The final set,
which was new in this phase of Galaxy~Zoo and designed for high-redshift targets, asked users to identify
objects dominated by clumpy morphologies. Further annotations for clumpy
galaxies included assessing the number, arrangement, relative brightness, and
location of the clumps within the galaxy. Finally, every classification of a
galaxy gave users the option of identifying ``odd'' features within the image;
these labels were for relatively rare $(\lesssim1\%)$ phenomena, including dust
lanes, gravitational lenses, and mergers.

@@

The approach used in GZH for correcting the weighted classifications for user
bias rests on the assumption that the \emph{amount} of bias is a function of
the apparent size and brightness of the image as seen on screen. This is
controlled by two types of parameters: \textbf{intrinsic} properties of the
galaxy itself, such as its physical diameter and luminosity, and
\textbf{extrinsic} properties, such as the distance (redshift) of the galaxy
and its relative orientation. There are likely other parameters that affect
user accuracy, such as the proximity of close companions \citep[``distraction
bias''; see][]{joh15} or bias as a function of individual user ability and
skill. The combination of all such parameters forms a high-dimensional space,
and it is not clear how to separate this into individual effects. Instead, the
method used here employs only two parameters intended to capture the bulk of
the change in bias: the $r^\prime$-band surface brightness ($\mu_r$; intrinsic)
and redshift ($z$; extrinsic).

@@

As with the HST images, each \ferengi{} subject has a varying number of users
answering the various questions in the hierarchical decision tree. Every user
answers the first question, {\it ``Is the galaxy smooth and rounded, with no
sign of a disk?''}; as such the vote fractions \fsmooth, \ffeatures, and
\fartifact{} are all computed with the minimum statistical error for any
question, with between 40 and 140~responses (see Section~\ref{sec:interface}).
The number of users to answer any subsequent question, however, is always equal
to or less than the number to answer the preceding question. The average number
of responses per task for fourth- or fifth-tier questions such as spiral arm
structure (Tasks 12--14) is only $4\pm4$ for the \ferengi{} sample; while this
is distribution is strongly bimodal (reflecting the true morphologies of
selected galaxies), the very low absolute numbers of votes introduce very high
variance when attempting to calculate a statistical correction.

In the \ferengi{} data, these numbers severely limit the amount of information
that can be extracted for the higher-tier questions. The debiasing technique
used (Section~\ref{ssec:zeta}) requires that at least 10~users answer each
question for a galaxy with $z_\mathrm{sim}=0.3$ \emph{and} the corresponding
image at higher redshift. This requirement is (by default) met by all galaxies
for the smooth/features question. However, this is often \emph{not} met for
questions beyond Task~01; on average. $60\%\pm24\%$ of the galaxies do not have
sufficient data to measure a correction, as compared to 2.0\% achieved for
Task~01 (Table~\ref{tbl:ferengi_corrections}). Of the remaining
\ferengi{} galaxies which did meet this requirement, polynomials were fit to
the data in each redshift-surface brightness bin, as was done for Task~01. The
goodness-of-fit was evaluated in each bin, for each Task, using a normalized
$\chi^2$ metric. The mean $\chi^2$ for all higher-order tasks was
$0.10\pm0.04$, significantly larger than the Task~01 value of $0.04$. Based on
this statistic, the data cannot not be modeled with sufficient confidence for
the higher order tasks. An example of the debiasing technique applied to
\fbar{} is shown in Section~\ref{sec:ferengi_bar}; visual inspection of
Figure~\ref{fig:f_vs_fbar} confirms that the points in each bin are scattered
to a degree at which no strong correlation is present.  For these reasons,
debiased vote fractions are only offered in the GZH catalog for Task~01
(smooth/features).

@@

Each image is listed under a unique project ID (eg,~AHZ000001); the actual
galaxy in the image is identified by the combination of the OBJNO and original
survey. For each of the 55~responses in the GZH decision tree, the following
classification data is provided: for each question, $\tt N_{votes}$ is the
number of users to answer that question. For each unique answer, $\tt fraction$
is the fraction of users to select that answer ($\rm N_{answer}/N_{votes}$, and
$\tt weighted$ is the weighted fraction, which takes into account user
consistency (Section~\ref{ssec:weighting}).

@@

In GZH, volunteers answer questions about a galaxy's morphology in a decision
tree format. With this structure, questions shown to a user are dependent on
their answers to the previous questions (Section~\ref{ssec:interface}).
Figure~\ref{fig:decisiontree} shows the possible paths available for answering
the questions offered in GZH. The colors represent the tier level of Tasks,
which indicate the number of previous Tasks the given Task is dependent upon.
The arm-number question, Task~11, is a 5$^\mathrm{th}$-tier Task, meaning that
whether this question is seen by a user is dependent on four Tasks preceding
it. In this case, it is only shown to volunteers who voted the galaxy was
featured/disk-shaped in Task~01, not clumpy in Task~12, not edge-on in Task~02,
and had spiral structure in Task~10.

To select galaxies of a morphological type identified with a particular Task, a
cut is placed on the vote fraction for that Task ($f_{\rm task}$), \emph{as
well as} the vote fractions for the Tasks preceding it, because of the
dependency induced by the decision-tree structure. For example, to select
barred galaxies, a cut may be placed on \fbar{} such that only galaxies where a
high fraction of votes for this task voted for the $bar,yes$ answer. This is
not the only necessary cut, however, since not all users answer this question;
only those who have previously selected ``features'' in Task~01, ``not clumpy''
in Task~12, and ``not edge-on'' in Task~02 will have the opportunity to vote on
the bar question, Task~03. To ensure that \fbar{} is well-sampled, cuts on all
previous tasks must be applied.

The flexibility of this catalog allows the user to set their own selection
criteria for vote fraction thresholds to create a morphologically pure sample.
Table~\ref{tbl:thresholds} provides baseline cuts for selecting galaxies of a
variety of morphologies. These thresholds are determined by a visual inspection
of subsamples. For each Task, at least 20 users are required to have voted on
the given question along with a cut on the vote fraction is applied to the
previous task. We visually analyzed subsamples of 50~galaxies meeting both
criteria, as well as a control sample of galaxies which had 20~users vote on
the task, but did not meet the threshold cut set for the previous task. The
threshold cut was adjusted and new subsamples were inspected until both the
original and control samples achieved $>80\%$ purity.

@@

Tasks 03, 04 and 05 have an additional possible pathway; as shown in
Figure~\ref{fig:decisiontree}, a user might also be shown this question if they
select ``featured/disk'' in Task~01, ``clumpy'' in Task~12, two or more clumps
in Task~16, and ``spiral arrangement'' in Task~15. After applying the
appropriate thresholds for this path, $< 0.5\%$ of the galaxies which have $\ge
20$ answers to these questions used this pathway to arrive at these Tasks.
Further, of these subjects, none were found to exhibit disk structure, although
the clumps within were arranged in a spiral pattern.

@@

GZH used both 5-epoch and 2-epoch sets of data to construct the GOODS set of
images. The 11,157 full depth 5-epoch images are used in the main catalog; the
classifications for the 7,464 shallow depth 2-epoch images are provided as a
supplementary table. This section analyzes the effect of image depth on the
ability of the GZ users to identify features or disk structure in the images.

>>>>

% Note: 250-word limit
We present the data release paper for the Galaxy Zoo: Hubble (GZH)
project.\footnote{This publication has been made possible by the participation
of more than 200,000~volunteers in the Galaxy Zoo project. Their contributions
are individually acknowledged at
\url{http://authors.galaxyzoo.org/authors.html}.} This is the third phase in a
large effort to measure reliable, detailed morphologies of galaxies by using
crowdsourced visual classifications of color composite images. Images in GZH
were selected from various \textit{Hubble Space Telescope} Legacy programs that
used the Advanced Camera for Surveys (AEGIS, COSMOS, GEMS, GOODS-N, and
GOODS-S), with filters selected to probe the rest-frame optical emission from
galaxies out to $z\simeq1$. The galaxies selected for GZH classifications go
down to magnitude limits of $m_{I814W}<23.5$ and have a median redshift of
$\langle z\rangle=0.9\pm0.6$, with a tail extending out to $z\simeq4$. The GZH
morphological data include measurements of both bulge- and disk-dominated
galaxies, details on spiral disk structure that relate to the Hubble type, as
well as clump identification and geometry. This paper also describes and
employs a novel method for calibrating morphological bias by using
artificially-redshifted galaxy images as a baseline. The GZH catalog contains
both raw and calibrated morphological vote fractions for $150,771$ galaxies,
providing a comprehensive data set suitable for large-scale studies of galaxy
evolution. 
% didn't the GOODS-N and GOODS-S second phases go deeper than 23.5 mag? am I misremembering? --BDS

@@

\subsection{Classifier weighting}\label{ssec:weighting}

The votes of individual classifiers in GZH were combined to
make a vote fraction for each response ($f_{response}$) to a question in the
classification tree. Votes were weighted \citep[in a method identical to
that described in][]{wil13} such that classifiers who frequently disagreed with
others ended up having very low weights. The weighting $(w)$ was 1
for the top 95\% of classifiers, as ranked by consistency. For the bottom 5\% of
classifiers, $w$ was designed to drop smoothly and was effectively zero for the
bottom 1\%.  Since this only affected a tiny percentage of
the classifiers (and an even smaller percentage of the classifications) the
overall effect on the GZH dataset was minimal. The method was effective,
however, at filtering out contributions from random or deliberately
malicious classifiers.

Classifications were aggregated and weighted only if the classifier was logged
into GZH under their username (which was encouraged, but not required for
participation). Classifications by participants who were not logged in were marked as
``Anonymous''; no weighting was applied, but the data were otherwise treated
identically in the data processing.

@@

\begin{figure*}
\center
\includegraphics[width=160mm]{figures/gzh_interface.png}
\caption{Screenshot of the GZH interface (\url{http:/zoo3.galaxyzoo.org}) at
the beginning of a classification, with the classifier ready to select an answer for
the first question in the decision tree.\label{fig:interface}}
\end{figure*}

Classifications for GZH were made using a web-based interface
(Figure~\ref{fig:interface}), similar in design to Galaxy~Zoo and Galaxy~Zoo~2.
The front-end runs on a Ruby~on~Rails framework with classifications stored in
a MySQL backend. Classifiers were shown a randomly-selected color composite image from
the GZH sample; the default showed the image with a black sky background,
although they had the option to invert the color palette if desired. The
questions and responses for morphology appeared on the right side of the image as
a panel, including both text and icons. There was no tutorial required for
participation, although classifiers could access an extensive ``Help'' section
containing example images and descriptive text for all the morphological
labels.

\begin{figure*}
\center
\includegraphics[width=\textwidth]{figures/gzh_decision_tree.pdf}
\caption{Flowchart of the questions presented to GZH classifiers, labeled with
the corresponding Task numbers. Tasks in the decision tree are color-coded by
tier level: Gray-colored Tasks are 1$^\mathrm{st}$-tier questions which are
asked in every classification. Tasks colored green, blue, purple, and pink
(respectively) are one, two, three, or four steps below branching points in the
decision tree.}
\label{fig:decisiontree}
\end{figure*}

The procedure for classifying an image in GZH followed a hierarchical decision
tree (Figure~\ref{fig:decisiontree}). Every classification began with the step of
identifying whether the object at the center of the image was a smooth galaxy,
a galaxy with a disk or other features, or a star/artifact. Subsequent
questions in the tree depended on the previous answer(s) given by the classifier; the
decision tree was designed so that every question relevant to
the morphology in the process of being identified was answered. Questions that were not
answered were implicitly assumed to be absent in the image --- for example, if
the classifier identified a galaxy as being smooth, they were not asked to count the
number of spiral arms. For every task, the classifier chose a single answer before
continuing to the next question; they also had the option to restart any
classification in-progress.

The GZH decision tree had four broad sets of morphologies. The first option
designated stars or image artifacts (the result of either bad data or incorrect
classification by the ACS pipeline as a galaxy); in this case, the
classification process ended and no further questions were asked. The second
set were smooth galaxies, intended to select ellipticals/early-types;
classifiers also measured the relative axial ratio (roundness) for these
galaxies. The third set was for disk/late-type galaxies, which measured the
features necessary to place a galaxy on the standard Hubble tuning fork (bars,
spiral arm, strength of the central bulge). The final set, which was new in
this phase of Galaxy~Zoo and designed for high-redshift targets, identified
objects dominated by clumpy morphologies. Further annotations for clumpy
galaxies included assessing the number, arrangement, relative brightness, and
location of the clumps within the galaxy. Finally, every classification of a
galaxy gave the option of identifying ``odd'' features within the image; these
labels were for relatively rare $(\lesssim1\%)$ phenomena, including dust
lanes, gravitational lenses, and mergers.

@@

The approach used in GZH for correcting the weighted classifications for redshift
bias rests on the assumption that the \emph{amount} of bias is a function of
the apparent size and brightness of the image as seen on screen. This is
controlled by two types of parameters: \textbf{intrinsic} properties of the
galaxy itself, such as its physical diameter and luminosity, and
\textbf{extrinsic} properties, such as the distance (redshift) of the galaxy
and its relative orientation. There may also be other parameters that affect
classifier accuracy, such as the proximity of close companions \citep[``distraction
bias''; see][]{joh15} or bias as a function of individual classifier ability and
skill. The combination of all such parameters forms a high-dimensional space,
and it is not clear how to separate this into individual effects. Instead, the
method used here employs only two parameters intended to capture the bulk of
the change in bias: the $r^\prime$-band surface brightness ($\mu_r$; intrinsic)
and redshift ($z$; extrinsic).

@@

As with the HST images, each \ferengi{} subject had a varying number of classifiers
answering the various questions in the hierarchical decision tree. Every classifier
answers the first question, {\it ``Is the galaxy smooth and rounded, with no
sign of a disk?''}; as such the vote fractions \fsmooth, \ffeatures, and
\fartifact{} are all computed with the minimum statistical error for any
question, with between 40 and 140~responses (see Section~\ref{sec:interface}).
The number of participants answering subsequent questions, however, is always equal
to or less than the number who answered the preceding question. The average number
of responses per task for fourth- or fifth-tier questions such as spiral arm
structure (Tasks 12--14) is only $4\pm4$ for the \ferengi{} sample; while this
is distribution is strongly bimodal (reflecting the true morphologies of
selected galaxies), the very low absolute numbers of votes introduce very high
variance when attempting to calculate a statistical correction.

In the \ferengi{} data, these numbers severely limit the amount of information
that can be extracted for the higher-tier questions. The debiasing technique
used (Section~\ref{ssec:zeta}) requires that at least 10~classifiers answer each
question for a galaxy with $z_\mathrm{sim}=0.3$ \emph{and} the corresponding
image at higher redshift. This requirement is (by default) met by all galaxies
for the smooth/features question. However, this is often \emph{not} met for
questions beyond Task~01; on average. $60\%\pm24\%$ of the galaxies do not have
sufficient data to measure a correction, as compared to 2.0\% achieved for
Task~01 (Table~\ref{tbl:ferengi_corrections}). Of the remaining
\ferengi{} galaxies which did meet this requirement, polynomials were fit to
the data in each redshift-surface brightness bin, as was done for Task~01. The
goodness-of-fit was evaluated in each bin, for each Task, using a normalized
$\chi^2$ metric. The mean $\chi^2$ for all higher-order tasks was
$0.10\pm0.04$, significantly larger than the Task~01 value of $0.04$. Based on
this statistic, the data cannot not be modeled with sufficient confidence for
the higher order tasks. An example of the debiasing technique applied to
\fbar{} is shown in Section~\ref{sec:ferengi_bar}; visual inspection of
Figure~\ref{fig:f_vs_fbar} confirms that the points in each bin are scattered
to a degree at which no strong correlation is present.  For these reasons,
debiased vote fractions are only offered in the GZH catalog for Task~01
(smooth/features).

@@

Each image is listed under a unique project ID (eg,~AHZ000001); the actual
galaxy in the image is identified by the combination of the OBJNO and original
survey. For each of the 55~responses in the GZH decision tree, the following
classification data is provided: for each question, $\tt N_{votes}$ is the
number of classifiers who answered that question. For each unique answer, $\tt fraction$
is the fraction of classifiers who selected that answer ($\rm N_{answer}/N_{votes}$, and
$\tt weighted$ is the weighted fraction, which takes into account overall
consistency (Section~\ref{ssec:weighting}).

@@

In GZH, participants answered questions about a galaxy's morphology in a decision
tree format. With this structure, questions shown to a classifier were dependent on
their answers to the previous questions (Section~\ref{ssec:interface}).
Figure~\ref{fig:decisiontree} shows the possible paths available for answering
the questions offered in GZH. The colors represent the tier level of Tasks,
which indicate the number of previous Tasks the given Task is dependent upon.
The arm-number question, Task~11, is a 5$^\mathrm{th}$-tier Task, meaning that
whether this question is seen by a classifier was dependent on four Tasks preceding
it. In this case, it as only shown to classifiers who identified the galaxy as
featured/disk-shaped in Task~01, not clumpy in Task~12, not edge-on in Task~02,
and having spiral structure in Task~10.

To select galaxies of a morphological type identified with a particular Task, a
cut is placed on the vote fraction for that Task ($f_{\rm task}$), \emph{as
well as} the vote fractions for the Tasks preceding it, because of the
dependency induced by the decision-tree structure. For example, to select
barred galaxies, a cut may be placed on \fbar{} such that only galaxies where a
high fraction of votes for this task voted for the $bar,yes$ answer. This is
not the only necessary cut, however, since not all classifiers answer this question;
only those who have previously selected ``features'' in Task~01, ``not clumpy''
in Task~12, and ``not edge-on'' in Task~02 will have the opportunity to vote on
the bar question, Task~03. To ensure that \fbar{} is well-sampled, cuts on all
previous tasks must be applied.

The flexibility of this catalog allows users to set their own selection
criteria for vote fraction thresholds to create a morphologically pure sample.
Table~\ref{tbl:thresholds} provides baseline cuts for selecting galaxies of a
variety of morphologies. These thresholds are determined by a visual inspection
of subsamples. For each Task, at least 20 classifiers are required to have voted on
the given question along with a cut on the vote fraction is applied to the
previous task. We visually analyzed subsamples of 50~galaxies meeting both
criteria, as well as a control sample of galaxies which had 20~classifiers vote on
the task, but did not meet the threshold cut set for the previous task. The
threshold cut was adjusted and new subsamples were inspected until both the
original and control samples achieved $>80\%$ purity.

@@

Tasks 03, 04 and 05 have an additional possible pathway; as shown in
Figure~\ref{fig:decisiontree}, a classifier might also be shown this question if they
select ``featured/disk'' in Task~01, ``clumpy'' in Task~12, two or more clumps
in Task~16, and ``spiral arrangement'' in Task~15. After applying the
appropriate thresholds for this path, $< 0.5\%$ of the galaxies which have $\ge
20$ answers to these questions used this pathway to arrive at these Tasks.
Further, of these subjects, none were found to exhibit disk structure, although
the clumps within were arranged in a spiral pattern.

@@

GZH used both 5-epoch and 2-epoch sets of data to construct the GOODS set of
images. The 11,157 full depth 5-epoch images are used in the main catalog; the
classifications for the 7,464 shallow depth 2-epoch images are provided as a
supplementary table. This section analyzes the effect of image depth on the
ability of the GZ classifiers to identify features or disk structure in the images.
